---
title: "Data_Exploration"
author: "Rebecca Sullivan"
date: "4/29/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if(! require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  here,
  lubridate
)


```

What are the recent different trends in weather patterns across these regions that can increase the need for emergency roadside services?  
   to help distribute resources accordingly in the 3 different areas
  What sorts of services:  Dead Batteries and Towing
  dead batteries can be triggered by temperature changes

```{r}

weather_1 <- read_csv(here::here("Data", "2938751.csv")) %>% 
  select (STATION, NAME, LATITUDE, LONGITUDE,  ELEVATION, DATE, PRCP, SNOW, TMAX, TMIN  )


head(weather_1)
tail(weather_1)
```

```{r}

weather_2 <- read_csv(here::here("Data", "2966446.csv")) %>% 
  select (STATION, NAME, LATITUDE, LONGITUDE,  ELEVATION, DATE, PRCP, SNOW, TMAX, TMIN  )



head(weather_2)
```


```{r}

weather_3 <- read_csv(here::here("Data", "2966466.csv")) %>% 
  select (STATION, NAME, LATITUDE, LONGITUDE,  ELEVATION, DATE, PRCP, SNOW, TMAX, TMIN  )



head(weather_3)
```




```{r}
weather <- union (weather_1, weather_2) %>% 
  union(weather_3) %>% 
  mutate (Location = NAME ,
          Year = year(DATE),
          Month = months(DATE),
          Month_num = month(DATE, label = FALSE) ,
          Quarter = quarters(DATE), 
          Daily_Temp_Change = TMAX - TMIN) 

head(weather)


```

```{r}
str(weather)
summary(weather)
```

```{r}
weather %>% 
  count(STATION, Location)
```



I requested weather data from 4 parts of the country.  I did find some metrics in separate files on the site, but not all have temperature and precipitation measurements.

The first station I found in region 4 with temperature and weather is GOULD 4 in CO.  I also requested data from SD, since that is data is closer to the region serviced.  Based on this exploration, I will only use one of those as the 4th area. 

```{r}
weather %>% 
  filter( STATION != 'USC00053446' ) %>% 
  ggplot(aes ( y = TMAX ,
               group = Location, 
               color = Location)) +
  geom_boxplot(na.rm = T )    +
  labs(
        y = "Maximum Daily Temperature in Degrees F" ,
    title = "Daily Temperature Variation 2019 - 2021" ,
    subtitle = "by Weather Station",
    caption = "Data from www.noaa.gov") +
    theme_gray()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )


  weather %>% 
  filter( STATION != 'USC00053446' ) %>% 
  ggplot( aes ( x =  TMAX )) +
  geom_histogram(na.rm = TRUE)+
  theme_bw()


```

```{r}
weather %>% 
  filter( STATION != 'USC00053446' ) %>% 
  ggplot(aes ( y = TMIN,
               group = Location, 
               color = Location)) +
  geom_boxplot(na.rm = T ) +
      labs(
        y = "Minimum Daily Temperature in Degrees F" ,
    title = "Daily Temperature Variation 2019 - 2021" ) +
    theme_gray()  


  weather %>% 
  filter( STATION != 'USC00053446' ) %>% 
  ggplot( aes ( x =  TMIN )) +
  geom_histogram(na.rm = TRUE)+
  theme_bw()



```
```{r}
weather %>% 
  filter( STATION != 'USC00053446' ) %>% 
  ggplot(aes ( y = Daily_Temp_Change,
               group = Location, 
               color = Location)) +
  geom_boxplot(na.rm = T ) +
  labs(
        y = "Daily Temperature Change in Degrees F" ,
    title = "Temperature Variation 2019 - 2021" ) +
    theme_gray()   
  

```



```{r}
weather %>% 
  count(Location, Year) 

```
```{r}
weather %>% 
  count(Location, Year, Month) 

```

Full data for 3 recent years from 3-4 different areas of the country
All in the footprint of customers that work serves

Keep only the 3 full years of measures, and add flags for below freezing and above 90
```{r}
 
  weather %>% 
  filter( Year %in% c(2019, 2020, 2021)) %>% 
  mutate( cold_flag = if_else(condition = TMIN <= 32, 
                                  true = T, 
                                  false = F, 
                              missing = NA)   ) %>%
    mutate( hot_flag = if_else(condition = TMAX >= 90, 
                                  true = T, 
                                  false = F, 
                              missing = NA) ) %>% 
  count(cold_flag, hot_flag) 

weather <-
  weather %>% 
  filter( Year %in% c(2019, 2020, 2021)) %>% 
  mutate( cold_flag = if_else(condition = TMIN <= 32, 
                                  true = T, 
                                  false = F, 
                              missing = NA)   ) %>%
    mutate( hot_flag = if_else(condition = TMAX >= 90, 
                                  true = T, 
                                  false = F, 
                              missing = NA) ) 

head(weather)
```
does not appear we have any extreme days that have a low less than 32 and a high above 90, at least not in this data. 

```{r}
weather %>% 
  group_by(Location, Year) %>% 
  summarise( avg_min_temp= mean(TMIN, na.rm = TRUE),
             min_temp = min(TMIN, na.rm = TRUE) ,
             avg_max_temp = mean(TMAX, na.rm = TRUE),
             max_temp = max(TMAX, na.rm = TRUE)) %>% 
  ungroup()



```


```{r}
weather %>% 
  group_by(Location, Year, Month) %>% 
  summarise( min_temp= mean(TMIN, na.rm = TRUE),
             max_temp = mean(TMAX, na.rm = TRUE) ) %>% 
  ungroup()



```
All stations
```{r}

   weather %>% 
  ggplot( aes ( x =  Daily_Temp_Change )) +
  geom_histogram(na.rm = TRUE)+
  theme_bw()



weather %>% 
  ggplot( aes ( x = DATE,
                y = Daily_Temp_Change )) +
  geom_point(na.rm = TRUE)+
  geom_smooth(na.rm = TRUE) +
  theme_gray()


   


```
one station
```{r}
   weather %>% 
  filter( STATION == "USW00013739") %>% 
  ggplot( aes ( x =  Daily_Temp_Change )) +
  geom_histogram(na.rm = TRUE)+
  theme_bw()


weather %>% 
  filter( STATION == "USW00013739") %>% 
  ggplot( aes ( x = DATE,
                y = Daily_Temp_Change )) +
  geom_point(na.rm = TRUE)+
  geom_smooth(na.rm = TRUE) +
  theme_gray()




```
as expected, there is not much difference throughout the year.  A little bit of a pattern, which appears to be seasonal.  Zoom in on 1 year to check that. 

```{r}
weather %>% 
  filter( STATION == "USW00013739" & Year == 2020) %>% 
  ggplot( aes ( x = DATE,
                y = Daily_Temp_Change )) +
  geom_point(na.rm = TRUE)+
  geom_smooth(na.rm = TRUE) +
  theme_gray()


   


```


```{r, warning=FALSE}
weather %>% 
  filter( STATION == "USW00013739") %>% 
  ggplot( aes ( x = DATE,
                y = TMAX )) +
  geom_point(na.rm = TRUE)+
  geom_smooth(na.rm = TRUE) +
  theme_gray()


```
TMAX is in cycles also, as expected.  Hot in summer and cold in winter. 

  
```{r}
weather %>% 
  filter( STATION == "USW00013739") %>% 
  filter( is.na(TMAX) == FALSE) %>% 
  ggplot( aes ( x = DATE,
                y = TMAX )) +
  geom_col() +
  theme_gray()
```
Pivot so can graph min and max for same day, in different colors
and stack.


```{r}
weather %>% 
  filter( STATION == "USW00013739") %>% 
  filter( is.na(TMAX) == FALSE & is.na(TMIN) == FALSE) %>% 
  pivot_longer(starts_with("TM"),
               names_to = "type" ) %>% 
  ggplot( aes ( x = DATE,
                y = value, 
                fill = type)) +
  geom_col(position = "stack") +
  labs(
            x = "Date" ,
        y = "Temperature in Degrees F" ,
    fill = "Daily Max or Min",
    title = "Temperature Variation by Date 2019 - 2021" ,
    subtitle = "4 Locations Combined",
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )
  
```
```{r}
graph_temps <- function ( df,
               Station_code ="USW00013739" )
{ 
  df %>% 
  filter( STATION == Station_code) %>% 
  filter( is.na(TMAX) == FALSE & is.na(TMIN) == FALSE) %>% 
  pivot_longer(starts_with("TM"),
               names_to = "type" ) %>% 
  ggplot( aes ( x = DATE,
                y = value, 
                fill = type)) +
  geom_col(position = "stack") +
  labs(
            x = "Date" ,
            y = "Temperature in Degrees F" ,
    fill = "Daily Max or Min",
    title = "Temperature Variation by Date 2019 - 2021" ,
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  ) +
    facet_grid(~Location)

}

```


 seemed hard to read with a facet, trying with a function to make each individually
```{r}


weather %>% 
graph_temps(Station_code ="USW00013739")


weather %>% 
graph_temps(Station_code ="USC00396427")

weather %>% 
graph_temps(Station_code ="USW00013967")


weather %>% 
graph_temps(Station_code ="USW00093812")
```
The high and low temperatures do seem to cycle with the seasons.  Not seeing anything in the past 3 years that shows highs and lows don;t follow the same general pattern.  But there are min temperatures below 32 and highs above 90, and those together contribute to batter wear. 



Look at the number of above 90 and below 32
Note different times of year that have more extremes in each region  (3)

```{r, eval=FALSE}
weather %>% 
  group_by(Location, Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x =Year ,
                 y=  num_hot_days,
                color = Location,
                fill = Location)) +
  geom_col(position = "dodge") +
  theme(legend.position = "bottom" ,
        legend.direction = "vertical")

```
 Oklahoma appears to have the most hot days, and has seen an increase in hot days in the recent 3 years. South Dakota has the fewest hot days of the areas studied. 
```{r}
weather %>% 
  group_by(Location, Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x =Year ,
                 y=  num_cold_days,
                color = Location,
                fill = Location)) +
  geom_col(position = "dodge") +
  theme(legend.position = "bottom" ,
        legend.direction = "vertical")

```
 South Dakota has the most cold days.  Philadelphia has the lowest number of cold days, at least in the past 3 years.

When were these extreme hot and cold days during the year? Look at the months

```{r}

weather %>% 
  group_by(Location, Year, Month, Month_num) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x =Month,
                 y=  num_cold_days,
                color = Location,
                fill = Location) ) +
  geom_col(position = "dodge", na.rm = TRUE) +
  theme(legend.position = "bottom" ,
        legend.direction = "vertical",
        axis.text.x = (element_text(angle = 45)))


```
```{r, message=FALSE}

weather %>% 
  group_by(Location, Year, Quarter ) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = Quarter,
                 y=  num_cold_days,
                color = Location,
                fill = Location) ) +
  geom_col(position = "dodge", na.rm = TRUE) +
  theme(legend.position = "bottom" ,
        legend.direction = "vertical",
        )





```

```{r}

weather %>% 
  group_by(Location, Year, Quarter ) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = Quarter,
                 y=  num_hot_days,
                color = Location,
                fill = Location) ) +
  geom_col(position = "dodge", na.rm = TRUE) +
  theme(legend.position = "bottom" ,
        legend.direction = "vertical",
        )

```




lag? change from 1 day to the next.  especially if 1 day is below 32 or above 90/  (5)


a 2 day Temp difference of 60 between 2 days will also put stress on a battery.  Especially when one of those days was below 32
```{r}

weather %>% 
  mutate(Temp_max_DoD = TMAX - lag(TMAX, default = 0),
         Temp_min_DoD = TMIN - lag(TMIN, default = 0),
         Temp_2day_diff = abs(TMAX - lag(TMIN)))

weather %>% 
  mutate( Temp_1Week_diff = abs(TMIN  - lag(TMAX, n = 7)),
          Temp_2day_diff = abs(TMIN - lag(TMAX , n=1)))
```
I tried a few ways to lag, and then one I think makes the most sense is to compare the minimum and maximum from 1 day apart and 1 week apart


```{r}
weather %>% 
  mutate( Temp_2day_diff = abs(TMAX - lag(TMIN, default = 0))) %>% 
  ggplot(aes( x = DATE,
              y =Temp_2day_diff, 
              color = Location)) +
  geom_point(na.rm = TRUE) +
  geom_hline( aes(yintercept = 60)) +
  geom_hline( aes(yintercept = 50)) 


```
Philadelphia did have 1 day where there was more than 60 degrees of change
South Dakota seems to have more that are at least more than 50 degrees of change


difference over 1 week

```{r}
weather %>% 
  mutate( Temp_1Week_diff = abs(TMAX - lag(TMIN, default = 0, n = 7))) %>% 
  ggplot(aes( x = DATE,
              y =Temp_1Week_diff, 
              color = Location)) +
  geom_point(na.rm = TRUE) +
  geom_hline( aes(yintercept = 60)) +
  geom_hline( aes(yintercept = 50)) 


```
When we look over a 1 week, Oklahoma had the most in recent years 



When are these extreme changes between multiple days, during the year?  Look at the month or quarter
```{r}
weather %>% 
  mutate( Temp_1Week_diff = abs(TMAX - lag(TMIN, default = 0, n = 7)),
           Temp_2day_diff = abs(TMAX - lag(TMIN, default = 0)) ) %>% 
  select( Location, Year, Quarter, Month, TMAX, TMIN, Temp_1Week_diff, Temp_2day_diff) %>% 
  filter( Temp_1Week_diff >= 50 | Temp_2day_diff >=50 ) %>% 
  ggplot( aes(x = Quarter,
              fill = Location)) +
  geom_bar(position = "dodge", na.rm = TRUE) +
  theme(legend.position = "bottom" ,
        legend.direction = "vertical"
        )


```







Is latitude, longitude or elevation more correlated with extreme temperatures??
(more data points would help, but finding weather stations with temperature readings was a challenge)

```{r}
#graph number of extremes




weather %>% 
  filter( is.na(TMAX) == FALSE ) %>% 
  group_by(STATION,LONGITUDE, Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = LONGITUDE,
                y = num_hot_days 
               )) +
  geom_point() +
  geom_smooth(na.rm = TRUE) +
  labs(
            x = "Longitude" ,
        y = "Days Above 90 Degrees" ,
    title = "Temperature from 2019 - 2021" ,
    subtitle = "4 Locations Combined",
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )




```

```{r}
weather %>% 
  filter( is.na(TMAX) == FALSE ) %>% 
  group_by(STATION,LATITUDE, Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = LATITUDE,
                y = num_hot_days 
               )) +
  geom_point() +
  geom_smooth(na.rm = TRUE) +
  labs(
            x = "Latitude" ,
        y = "Days Above 90 Degrees" ,
    title = "Temperature from 2019 - 2021" ,
    subtitle = "4 Locations Combined",
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )
```


```{r}
weather %>% 
  filter( is.na(TMAX) == FALSE ) %>% 
  group_by(STATION, ELEVATION , Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = ELEVATION,
                y = num_hot_days 
               )) +
  geom_point() +
  geom_smooth(na.rm = TRUE) +
  labs(
            x = "Elevation" ,
        y = "Days Above 90 Degrees" ,
    title = "Temperature from 2019 - 2021" ,
    subtitle = "4 Locations Combined",
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )
```

```{r}
weather %>% 
  filter( is.na(TMAX) == FALSE ) %>% 
  group_by(STATION,LATITUDE, Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = LATITUDE,
                y = num_cold_days 
               )) +
  geom_point() +
  geom_smooth(na.rm = TRUE) +
  labs(
            x = "Latitude" ,
        y = "Days Below 32 Degrees" ,
    title = "Temperature from 2019 - 2021" ,
    subtitle = "4 Locations Combined",
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )
```
```{r, warning=FALSE}


weather %>% 
  filter( is.na(TMAX) == FALSE ) %>% 
  group_by(STATION,LONGITUDE, Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = LONGITUDE,
                y = num_cold_days 
               )) +
  geom_point() +
  geom_smooth(na.rm = TRUE) +
  labs(
            x = "Longitude" ,
        y = "Days Below 32 Degrees" ,
    title = "Temperature from 2019 - 2021" ,
    subtitle = "4 Locations Combined",
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )



```

```{r, warning=FALSE}
weather %>% 
  filter( is.na(TMAX) == FALSE ) %>% 
  group_by(STATION, ELEVATION , Year) %>% 
  summarise( num_hot_days = sum(as.numeric(hot_flag)),
             num_cold_days = sum(as.numeric(cold_flag))) %>% 
  ungroup() %>% 
  ggplot( aes ( x = ELEVATION,
                y = num_cold_days 
               )) +
  geom_point() +
  geom_smooth( na.rm = TRUE) +
  labs(
            x = "Elevation" ,
        y = "Days Below 32 Degrees" ,
    title = "Temperature from 2019 - 2021" ,
    subtitle = "4 Locations Combined",
    caption = "Data from www.noaa.gov") +
    theme_light()  +
    theme(axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "italic")  )
```




is there enough climate change to reconsider our strategy?
(maybe look at the same stations in 1990s or 1980s)  => scatter plot
are high or low temps actually correlated to the year?
number of extremes correlated??




```{r}
weather %>% 
  filter( STATION != 'USC00053446' ) %>% 
  group_by(Location ) %>% 
  ggplot(aes ( x = DATE,
                y = SNOW,
               group = Location, 
               color = Location)) +
  geom_boxplot(na.rm = T )


```



```{r}
weather %>% 
  filter( STATION != 'USC00053446' ) %>% 
  group_by(Location ) %>% 
  ggplot(aes ( x = DATE,
                y = PRCP,
               group = Location, 
               color = Location)) +
  geom_boxplot(na.rm = T )

```
